---
title: "Insert the Chapter or Lab Report name here"
author: "Insert your name(s) here"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r load-packages, message=FALSE}
library(tidyverse)
```

# set working directory for all chunks in this file (default working directory is wherever Rmd file is)
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'set/working/directory/here')
```

### Question or Discussion Question or Challenge or Stop/Think/Do etc. 
### Outside a code chunk, ### indicates a header

Insert human-readable text here. This is where you can introduce a code chunk (following this) or sometimes the answer will not need code.

```{r}
# Insert code for Question 1 here
# Inside a code chunk, a # indicates a comment and will not run
```

You can add more text for Question 1 here if you want to. Make sure you show all your work/code. 


### Question or Discussion Question or Challenge or Stop/Think/Do etc. 

Insert human-readable text here. 

```{r}
# Insert code for Question 2 here
```

You can add more text here. 

### Repeat these blocks of headers (with ###), text, and code chunks for each Question on the Lab Manual. To add another code chunk, you can either copy paste a previous empty one, or by pressing the green +C button in the top right corner. 

When you are done (or just to check how it is looking), you can click the Knit button at the top. Default is usually to Knit to html---this is how you can turn your project into a website! To submit your Lab Report, instead Knit to PDF using the triangle drop-down and submit on Canvas. 
